Finally Tagless, Partially Evaluated

### 2007-06-12 tagless/tagless-final.tex
down to 19 pages. There is some work to be done in Sec 4.

\begin{quote}
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm\citep{reynolds-definitional}
\end{quote}

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage.
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\cite{metaocaml}) and standard Haskell. The complete code is available%
\footnote{\oleg{put URL!}.}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

@@

\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{variations}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above%
%aplas: (\S\ref{selfinterp})
.  Self\hyp interpretation turned out
  harder than expected%
%aplas
\footnote{Due to the severe lack of space, we cannot describe
self-interpretation and refer the reader to the accompanying code.}%
.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\cite{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation%
%aplas-specific
, Fig.~\ref{fig:interpreter-R}(a)
.
% fig:interpreter-R part (a) was here
%
\begin{figure}
(a) \begin{code2}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

  let int  (x:int)  = x
  let bool (b:bool) = b

  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self

  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code2}

(b) \begin{code2}
module EXR = EX(R)
\end{code2}

(c) \begin{code2}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code2}
\caption{The tagless interpreter \texttt{R}.}
\label{fig:interpreter-R}
\end{figure}
%
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
%aplas-specific
, Fig.~\ref{fig:interpreter-R}(b)
.
% fig:interpreter-R part (b) was here
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
%aplas-specific
 Fig.~\ref{fig:interpreter-R}(c)
.
% fig:interpreter-R part (b) was here
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
These propositions follow immediately from the soundness of the
metalanguage's type system.

@@

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the simple compiler~|C| below%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(a)
.
% fig:interpreter-C part (a) was here
%
\begin{figure}[t]
% this code does not have the 'sv parameter. It shows up later.
(a) \begin{code2}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code

  let int (x:int)   = .<x>.
  let bool (b:bool) = .<b>.

  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
end
\end{code2}

@@

This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and does not
  call the interpreter, unlike the recursive calls to |eval0| and
  |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.

We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
  (TH). The output of our compiler will then be assuredly type-correct TH.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
%aplas
\begin{comment}
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
\end{comment}
The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.

@@

To enable these simplifications, we delay binding-time analysis (BTA)
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\begin{code}
data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
\end{code}
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.

Thus typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type.  The connection with CPS
is not an accident, as we shall see in~\S\ref{S:CPS}.

@@

Two common ways to provide typecase in Haskell are
GADTs and type-class functional dependencies
\cite{oliveira-typecase}.  These
methods are equivalent, and here we show the GADT way; |incope1.hs|
in the accompanying source code shows the latter.

\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr

@@

  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
\label{fig:ocaml}
\end{figure*}

We introduce a GADT with four data constructors.
\begin{code}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b)
  E  :: C t -> P t
\end{code}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|. As before, we define a
function to extract a future-stage computation from a value of type |P t|.
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}
The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
case uses the method |lam| of the |C| interpreter to compile~|f|.

This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in |app|
above: the type |P t| has four constructors, of which pattern\hyp
matching in |app| uses only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
Indeed the metalanguage implementation could reason thus, 
but it may not; GHC for one issues warnings%
\footnote{If we use inductive families (as those in Coq) or logical
  frameworks with canonical forms, we can prove the pattern-matching
  to be exhaustive. The coverage checker of Twelf does such proof.}. 
Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but never fail in reality. The goal is to make this safety
\emph{syntactically} apparent.

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code}
repr_pe Int      = Int
repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.

\begin{figure}
\begin{floatrule}
\begin{code}
module P = struct
  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

@@

  let fix f = 
    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
    in let rec self = function
       | {st = Some _} as e -> app (f (lam self)) e
       | e -> pdyn (C.app fdyn (abstr e))
       in {st = Some self; dy = fdyn}
end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
\label{fig:pe}
\end{figure}

The interpreters |R|, |L| and~|C| above only use the old
type arguments |'c| and~|'dv|, which are treated by the new signature
in the same way.  Hence all that needs to change in these interpreters
to match the new signature is to add a phantom type
argument~|'sv| to~|repr|.
For example, the compiler |C| now begins
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
In contrast, the partial evaluator~|P| relies on the type argument |'sv|.

@@

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, unfold |fix| rather than residualize whenever the
argument is static.  In the accompanying source code is a conservative
alternative |P.fix| that unrolls recursion only once, then residualizes.
Many sophisticated approaches have been developed to make this trade-off
\cite{jones-partial}, but this issue is orthogonal to our presentation.
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\cite{SwadiTahaKiselyovPasalic2006}.

Given this implementation of~|P|, our running example
\begin{code}
let module E = EX(P) in E.test1 ()
\end{code}
evaluates to
\begin{code}
{P.st = Some true; P.dy = .<true>.}
\end{code}
of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
a $\beta$-reduction has been statically performed to yield |true|.  More
interestingly, whereas |testpowfix7| compiles to a code value with many
$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
\begin{code}
let module E = EX(P) in E.testpowfix7
\end{code}
gives the desired result
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 *
                    (x_1 * (x_1 * x_1)))))>.}
\end{code}

Unlike the GADT approach in~\S\ref{S:PE-GADT}, all pattern\hyp matching
in~|P| is \emph{syntactically} exhaustive, so it is patent to the metalanguage
implementation that |P| never gets stuck.  Further, all pattern\hyp matching occurs
during partial evaluation, only to check if a value is known statically,
never what type it has.  In other words, our partial evaluator tags
phases (with |Some| and |None|) but not object types.

>>>>

\begin{quote}
\small
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm\citep{reynolds-definitional}
\end{quote}

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage
%aplas
\footnote{Due to the most severe lack of space, we cannot describe
these features and refer the reader to the accompanying code.}%
.
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\cite{metaocaml}) and standard Haskell. The complete code is available
\url{\oleg{put URL!}.}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

@@

\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles%
% aplas
%   (\S\ref{variations}).
.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation%
% aplas: commented
%  and CPS transformation (\S\ref{PE}).
.
\item We point a clear way to extend the object language with more features
    such as state%
%aplas (\S\ref{state})%
%aplas
\footnotemark[\value{footnote}]
.
\item We describe an approach to self\hyp interpretation compatible with the
  above%
%aplas: (\S\ref{selfinterp})
.  Self\hyp interpretation turned out
  harder than expected%
%aplas
\footnotemark[\value{footnote}]
.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\cite{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation
\begin{code3}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
%aplas-specific
: \texttt{module EXR = EX(R)}%
.
%% \begin{code}
%% module EXR = EX(R)
%% \end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
% aplas: commented out
\begin{comment}
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
\end{comment}
%aplas: the following line is for aplas only
It is obvious to the compiler that
%
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
These propositions follow immediately from the soundness of the
metalanguage's type system.

@@

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the simple compiler~|C| below%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(a)
.
% fig:interpreter-C part (a) was here
%
\begin{figure}[t]
% this code does not have the 'sv parameter. It shows up later.
(a) \begin{code2}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code
  let int (x:int)   = .<x>.           let bool (b:bool) = .<b>.
  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2     = .<.~e1 + .~e2>. let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>. end
\end{code2}

@@

This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and does not
  call the interpreter, unlike the recursive calls to |eval0| and
  |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.
%aplas: the line is for aplas. The original paragraph commented out
The accompanying code shows the corresponding Haskell implemention. 
\begin{comment}
We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
  (TH). The output of our compiler will then be assuredly type-correct TH.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.
\end{comment}

@@

To enable these simplifications, we delay binding-time analysis (BTA)
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\begin{code3}
data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
\end{code3}
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
%aplas-simplification
Thus typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type.
\begin{comment}

Thus typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type.  The connection with CPS
is not an accident, as we shall see in~\S\ref{S:CPS}.
\end{comment}

@@

Two common ways to provide typecase in Haskell are
GADTs and type-class functional dependencies
\cite{oliveira-typecase}.  These
methods are equivalent, and here we show the GADT way; |incope1.hs|
in the accompanying source code shows the latter.
%aplas: the original text is in comments below. Here's a greatly
% abbreaviated text
We introduce a GADT with four data constructors.
\begin{code3}
data P t where
  VI :: Int  -> P Int;                VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b);   E  :: C t -> P t
\end{code3}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|. As before, we define a
function |abstr| to extract a future-stage computation from a value of 
type |P t|. The implementations of |int|, |bool|, and |add| are like
in~\S\ref{S:PE-problem}.
The interpretation of |lam f = VF f| just wraps the HOAS function |f|. 
We can always compile |f| to a code value,
but we delay it to apply |f| to concrete arguments. The interpretation of
\begin{code3}
app (VF f) ea = f ea;  app (E f)  ea = E (app f (abstr ea))
\end{code3}
checks to see if |ef| is such a delayed
HOAS function |VF f|. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
dynamic value |E f|, we residualize.
This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in
|app|. Although Coq or logical
frameworks with canonical forms could prove pattern-matching
exhaustive, GHC cannot and issues warnings. Recall, our goal is to
make this exhaustiveness \emph{syntactically} apparent.

\begin{comment}
We introduce a GADT with four data constructors.
\begin{code3}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b)
  E  :: C t -> P t
\end{code3}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|. As before, we define a
function to extract a future-stage computation from a value of type |P t|.
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}
The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
case uses the method |lam| of the |C| interpreter to compile~|f|.

This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in |app|
above: the type |P t| has four constructors, of which pattern\hyp
matching in |app| uses only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
Indeed the metalanguage implementation could reason thus, 
but it may not; GHC for one issues warnings%
\footnote{If we use inductive families (as those in Coq) or logical
  frameworks with canonical forms, we can prove the pattern-matching
  to be exhaustive. The coverage checker of Twelf does such proof.}. 
Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but never fail in reality. The goal is to make this safety
\emph{syntactically} apparent.
\end{comment}

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
%aplas: inlining
\texttt{data P t = P (Maybe (repr\_pe t)) (C t)}
\begin{comment}
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
\end{comment}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code3}
repr_pe Int      = Int          repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code3}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
% aplas: abbreviated figure
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int : int  -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x->'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation (\texttt{bool}, \texttt{mul},
\texttt{leq} are elided)}
\label{fig:ocaml}
\end{figure*}

\begin{comment}
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr

@@

  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
\label{fig:ocaml}
\end{figure*}
\end{comment}

% aplas: abberviated figure here. The original is in teh comments
% right down below
\begin{figure}
\begin{floatrule}
\begin{code}
module P = struct
  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x    let pdyn x = {st = None; dy = x}
  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}

  let app ef ea = match ef with {st = Some f} -> f ea
                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (\texttt{bool}, \texttt{mul},
  \texttt{leq} and \texttt{fix} are elided)}
\label{fig:pe}
\end{figure}

\begin{comment}
\begin{figure}
\begin{floatrule}
\begin{code}
module P = struct
  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

@@

  let fix f = 
    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
    in let rec self = function
       | {st = Some _} as e -> app (f (lam self)) e
       | e -> pdyn (C.app fdyn (abstr e))
       in {st = Some self; dy = fdyn}
end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
\label{fig:pe}
\end{figure}
\end{comment}

The interpreters |R|, |L| and~|C| above only use the old
type arguments |'c| and~|'dv|, which are treated by the new signature
in the same way.  Hence all that needs to change in these interpreters
to match the new signature is to add a phantom type
argument~|'sv| to~|repr|.
For example, the compiler |C| now begins \texttt{module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code} with the rest the same.
In contrast, the partial evaluator~|P| relies on the type argument |'sv|.

%aplas: here's the original text
\begin{comment}
For example, the compiler |C| now begins
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
\end{comment}

@@

%aplas: abbreviated paragraph. The original is commented below
Any partial evaluator must decide how much to unfold recursion. The code in
Figure~\ref{fig:pe} na\"\i vely unfolds |fix| whenever the
argument is static.  In the accompanying source code is a conservative
alternative |P.fix| that unrolls recursion only once, then residualizes.
Many sophisticated approaches have been developed for this trade-off
\cite{jones-partial}, but this issue is orthogonal to our presentation.
\begin{comment}
Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, unfold |fix| rather than residualize whenever the
argument is static.  In the accompanying source code is a conservative
alternative |P.fix| that unrolls recursion only once, then residualizes.
Many sophisticated approaches have been developed to make this trade-off
\cite{jones-partial}, but this issue is orthogonal to our presentation.
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\cite{SwadiTahaKiselyovPasalic2006}.
\end{comment}
